---
title: "MOVIE RECOMMENDATION SYSTEM"
author: "Parth Joshi
date: "27/04/2021"
output: word_document
---
**Aim:**R for data Science (Mini Project) 
•	 Extracting data from a large Dataset 
•	 Exploratory Analysis
•	 Use Mining Algorithm 
  Visualizations & interpretation of results.
  
**Lab Outcome No.:**2.ITL804.6

**Lab Outcome:**Apply the knowledge of R gained to data Analytics for real life applications.

**Problem Statement:*Recommendation systems play an important role in e-commerce and online streaming services, such as Netflix, YouTube, and Amazon. Making the right recommendation for the next product, music or movie increases user retention and satisfaction, leading to sales and profit growth. A recommendation system is a platform that provides its users with various contents based on their preferences and likings. The principal objective of this R programming project is to build a recommendation engine that recommends movies to users based on certain selected features from the dataset. The recommendation process can be fine-tuned by choosing the appropriate algorithm that fits the dataset and is less prone to error. The selection of the dataset should be based on users, user rating, title and it should have sufficiently large records of data for the 
given feature.


**Exploratory Analysis:**We are going to use the MovieLens Dataset which contains files such as movies.csv and ratings.csv file. The data consists of 105339 ratings applied over 10329 movies.
As mentioned the dataset consists of two csv files.
The movies.csv consists of 3 columns:
1.movieId: It is the unique id given to each movie.
2.title: This column gives you the title of the movie.
3.genre: This gives you the genre of the movie to which it belongs. We have 18 unique genre and each movie can belong to multiple genre hence they are separated by bar (|).
The ratings.csv consists of 4 columns:
1.userId:It is the unique id given to each user.
2.movieId:It is the unique id given to each movie. It is same as in movies.csv
3.ratings:Users have the option to choose a rating value from 0.5 to 5.0, totaling 10 possible values. This is unusual scale, so most movies get a rounded value rating.
4.timestamp:Timestamp are represented in seconds.


*Movie Recommendation System:* The main goal of this machine learning project is to build a recommendation engine that recommends movies to users. We will be developing an Item Based Collaborative Filter. Here we will calculate similarities between different items in the dataset by using one of a number of similarity measures, and then these similarity values are used to predict ratings for user-item pairs not present in the dataset.


**Code and Output:**

```{r}
#importing libraries
library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)

#reading dataset
setwd("D:\SEM 8\Rpl\R Project")

movie_data <- read.csv("movies.csv",stringsAsFactors=FALSE)
rating_data <- read.csv("ratings.csv")
class(rating_data$rating)

str(movie_data)

summary(movie_data)
head(movie_data)

summary(rating_data)
head(rating_data)
```

One-hot encoding is the process of converting a categorical variable with multiple categories into multiple variables, each with a value of 1 or 0. Used when there are movies which fall into multiple categories.
```{r}
###Converting categorical features to numeric
#One Hot Coding
movie_genre <- as.data.frame(movie_data$genres, stringsAsFactors=FALSE)

library(data.table)

movie_genre2 <- as.data.frame(tstrsplit(movie_genre[,1], '[|]', type.convert=TRUE), stringsAsFactors=FALSE)
colnames(movie_genre2) <- c(1:10)

list_genre <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")

genre_mat1 <- matrix(0,10330,18)
genre_mat1[1,] <- list_genre
colnames(genre_mat1) <- list_genre

for (index in 1:nrow(movie_genre2)) {
  for (col in 1:ncol(movie_genre2)) {
    gen_col = which(genre_mat1[1,] == movie_genre2[index,col])
    genre_mat1[index+1,gen_col] <- 1
  }
}

genre_mat2 <- as.data.frame(genre_mat1[-1,], stringsAsFactors=FALSE) #remove first row, which was the genre list

for (col in 1:ncol(genre_mat2)) {
  genre_mat2[,col] <- as.integer(genre_mat2[,col]) #convert from characters to integers
}

str(genre_mat2)

#Search Matrix
SearchMatrix <- cbind(movie_data[,1:2], genre_mat2[])
head(SearchMatrix)
```

For our movie recommendation system to make sense of our ratings through recommenderlabs,we have to convert our matrix into a sparse matrix one. This new matrix is of the class ‘realRatingMatrix’. This is performed as follows.
```{r}
#sparse matrix
ratingMatrix <- dcast(rating_data, userId~movieId, value.var = "rating", na.rm=FALSE)
ratingMatrix <- as.matrix(ratingMatrix[,-1]) #remove userIds
#Convert rating matrix into a recommenderlab sparse matrix
ratingMatrix <- as(ratingMatrix, "realRatingMatrix")
ratingMatrix
image(ratingMatrix[1:10,1:10])

#Viewing various options for building recommendation systems for movies.
recommendation_model <- recommenderRegistry$get_entries(dataType = "realRatingMatrix")
names(recommendation_model)

#Get the description of recommendation names.
lapply(recommendation_model, "[[", "description")

#In this project We will be implementing Item Based Collaborative Filtering model.Below are its parameters.
recommendation_model$IBCF_realRatingMatrix$parameters
```

Recommending movies is dependent on creating a relationship of similarity between the two users. With the help of recommenderlab, we can compute similarities using various operators like cosine, pearson etc.
```{r}
#Similarity Matrix
similarity_mat <- similarity(ratingMatrix[1:4, ],
                             method = "cosine",
                             which = "users")
as.matrix(similarity_mat)
image(as.matrix(similarity_mat), main = "User's Similarities")

movie_similarity <- similarity(ratingMatrix[, 1:4], 
                               method = "cosine", 
                               which = "items")
as.matrix(movie_similarity)
image(as.matrix(movie_similarity), main = "Movies similarity")

rating_values <- as.vector(ratingMatrix@data)
# extracting unique ratings
unique(rating_values) 

#create a table of ratings that will display the most unique ratings.
Table_of_Ratings <- table(rating_values) # creating a count of movie ratings
Table_of_Ratings
as.matrix(Table_of_Ratings)

library(ggplot2)
movie_views <- colCounts(ratingMatrix) # count views for each movie

table_views <- data.frame(movie = names(movie_views),
                          views = movie_views) # create dataframe of views
table_views <- table_views[order(table_views$views,
                                 decreasing = TRUE), ] # sort by number of views

table_views$title <- NA

for (index in 1:10325){
  table_views[index,3] <- as.character(subset(movie_data,
                                              movie_data$movieId == table_views[index,1])$title)
}
table_views[1:6,]
```

**Visualization:**
```{r}
#Visualize a bar plot for the total number of views of the top films.
ggplot(table_views[1:6, ], aes(x = title, y = views)) +
  geom_bar(stat="identity", fill = 'steelblue') +
  geom_text(aes(label=views), vjust=-0.3, size=3.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Total Views of the Top Films")
#From the bar-plot, we observe that Pulp Fiction is the most-watched film followed by Forrest Gump.

#Heat map
image(ratingMatrix[1:20, 1:25], axes = FALSE, main = "Heatmap of the first 20 rows and 25 columns")
```

Conduct **data preparation** in the following three steps –
 Selecting useful data.
 Normalizing data.
 Binarizing the data.
For finding useful data in our dataset, we have set the threshold for the minimum number of users who have rated a film as 50. This is also same for minimum number of views that are per film. This way, we have filtered a list of watched films from least-watched ones.
```{r}
movie_ratings <- ratingMatrix[rowCounts(ratingMatrix) > 50,
                              colCounts(ratingMatrix) > 50]
movie_ratings
image(movie_ratings)
print(movie_ratings)


minimum_movies<- quantile(rowCounts(movie_ratings), 0.98)
minimum_users <- quantile(colCounts(movie_ratings), 0.98)
image(movie_ratings[rowCounts(movie_ratings) > minimum_movies,
                    colCounts(movie_ratings) > minimum_users],
      main = "Heatmap of the top users and movies")

#Visualize the distribution of the average ratings per user.
average_ratings <- rowMeans(movie_ratings)
head(average_ratings)
qplot(average_ratings, fill=I("steelblue"), col=I("red")) +
  ggtitle("Distribution of the average rating per user")
```

Data Normalization:
In the case of some users, there can be high ratings or low ratings provided to all of the watched films. This will act as a bias while implementing our model. In order to remove this, we normalize our data. 
```{r}
normalized_ratings <- normalize(movie_ratings)

sum(rowMeans(normalized_ratings) > 0.00001)
image(normalized_ratings[rowCounts(normalized_ratings) > minimum_movies,
                         colCounts(normalized_ratings) > minimum_users],
      main = "Normalized Ratings of the Top Users")
```

Performing Data Binarization:
In the final step of our data preparation in this data science project, we will binarize our data. Binarizing the data means that we have two discrete values 1 and 0, which will allow our recommendation systems to work more efficiently. We will define a matrix that will consist of 1 if the rating is above 3 and otherwise it will be 0.
```{r}
binary_minimum_movies <- quantile(rowCounts(movie_ratings), 0.95)
binary_minimum_users <- quantile(colCounts(movie_ratings), 0.95)
#movies_watched <- binarize(movie_ratings, minRating = 1)
good_rated_films <- binarize(movie_ratings, minRating = 3)
image(good_rated_films[rowCounts(movie_ratings) > binary_minimum_movies,
                       colCounts(movie_ratings) > binary_minimum_users],
      main = "Heatmap of the top users and movies")
```

**Collaborative Filtering System**
In this section of project, we will develop Item Based Collaborative Filtering System. This type of collaborative filtering finds similarity in the items based on the people’s ratings of them. The algorithm first builds a similar-items table of the customers who have purchased them into a combination of similar items. This is then fed into the recommendation system.
```{r}
#Splitting the data into test and train
sampled_data<- sample(x = c(TRUE, FALSE),
                      size = nrow(movie_ratings),
                      replace = TRUE,
                      prob = c(0.8, 0.2))

training_data <- movie_ratings[sampled_data, ]
testing_data <- movie_ratings[!sampled_data, ]
image(training_data)


#Explore the various parameters of our Item Based Collaborative Filter.
recommendation_system <- recommenderRegistry$get_entries(dataType ="realRatingMatrix")
recommendation_system$IBCF_realRatingMatrix$parameters

recommen_model <- Recommender(data = training_data,
                              method = "IBCF",
                              parameter = list(k = 30))
recommen_model
class(recommen_model)
```

Using the getModel() function, we will retrieve the recommen_model. We will then find the class and dimensions of our similarity matrix that is contained within model_info. Finally, we will generate a heatmap, that will contain the top 20 items and visualize the similarity shared between them.
```{r}
model_info <- getModel(recommen_model)
class(model_info$sim)
dim(model_info$sim)
top_items <- 20
image(model_info$sim[1:top_items, 1:top_items],
      main = "Heatmap of the first 20 rows and columns")

#Sum of rows and columns with the similarity of the objects above 0.
sum_rows <- rowSums(model_info$sim > 0)
table(sum_rows)
sum_cols <- colSums(model_info$sim > 0)
#Visualize the sum of columns through a distribution
qplot(sum_cols, fill=I("steelblue"), col=I("red"))+ ggtitle("Distribution of the column count")

top_recommendations <- 10 #The number of items to recommend to each user
predicted_recommendations <- predict(object = recommen_model,
                                     newdata = testing_data,
                                     n = top_recommendations)
predicted_recommendations

user1 <- predicted_recommendations@items[[1]] # recommendation for the first user
movies_user1 <- predicted_recommendations@itemLabels[user1]
movies_user2 <- movies_user1
for (index in 1:10){
  movies_user2[index] <- as.character(subset(movie_data,
                                             movie_data$movieId == movies_user1[index])$title)
}
movies_user2

recommendation_matrix <- sapply(predicted_recommendations@items,
                                function(x){ as.integer(colnames(movie_ratings)[x]) }) # matrix with the recommendations for each user

#dim(recc_matrix)
recommendation_matrix[,1:4]

number_of_items <- factor(table(recommendation_matrix))
chart_title <- "Distribution of the Number of Items for IBCF"
qplot(number_of_items, fill=I("steelblue"), col=I("red")) + ggtitle(chart_title)

number_of_items_sorted <- sort(number_of_items, decreasing = TRUE)
number_of_items_top <- head(number_of_items_sorted, n = 4)
table_top <- data.frame(as.integer(names(number_of_items_top)),
                        number_of_items_top)
for(i in 1:4) {
  table_top[i,1] <- as.character(subset(movie_data,
                                        movie_data$movieId == table_top[i,1])$title)
}
colnames(table_top) <- c("Movie Title", "No. of Items")
head(table_top)


user2 <- predicted_recommendations@items[[2]] # recommendation for the first user
movies_user2 <- predicted_recommendations@itemLabels[user2]
movies_user3 <- movies_user2
for (index in 1:10){
  movies_user3[index] <- as.character(subset(movie_data,
                                             movie_data$movieId == movies_user2[index])$title)
}
movies_user3

```
